---
title: "Homework 2"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center")
library(tidyverse)
```

\bigskip

1. Recall `iris` data set. Find the maximum value of the sum `Sepal.Length` $+$ `Sepal.Width`. To which species does the sample with the maximum value you found above belong? _Hint_: you might want to start with defining a new variable `Sepal.Sum` using `mutate()`, for example. 

```{r, eval = F}
iris %>%
  mutate(sum = Sepal.Length + Sepal.Width) %>%
  arrange(desc(sum)) %>%
  slice(1) %>%
  select(sum, Species)
```

2. Create the following plot using `iris` data set. Note that `setosa` species do not show up. You do NOT need to adjust size or width/height ratio of the plot.

```{r, out.width = "50%"}
iris %>% filter(Species != "setosa") %>%
  ggplot(aes(Petal.Length, Petal.Width, color = Species)) +
  theme(aspect.ratio = 1) +
  geom_point()
```

3. Create the following plot using `diamonds` data set. 

```{r, fig.height = 5, out.width = "80%"}
ggplot(diamonds, aes(color, price)) +
  geom_boxplot() +
  facet_wrap(~cut)
```

\newpage

4. Determine $c$ so that the probability $0\leq Z \leq c$ is $0.45$, where $Z$ follows the standard normal distribution $N(0,1^2)$.

```{r, eval = F}
qnorm(0.95)
```

5. Recall the `heights.txt` data we used in class with 1375 mom/daughter height pairs:

```{r, echo = T}
ht <- read.table(file = "heights.txt", header = T, sep = " ")
```

Assuming that UK women's heights follow a normal distribution $N(\mu, \sigma^2)$, compute the 95% confidence interval of $\mu$ based on the mom's height data. 

```{r, eval = F}
t.test(ht$Mheight)
```


