---
title: "Unit 1"
author: "Jung-Jin Lee"
date: "Last modified: `r format(Sys.Date(), '%b %d, %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "Stan.css"] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, message = F, echo = F, warning = F}
library(knitr) 
opts_chunk$set(
  tidy = FALSE,
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 3,
  fig.height = 3
  )
```

## List of packages used

```{r}
library(tidyverse)
library(broom)
library(gt)
library(rstan)
# the following are recommended in startup messages
options(mc.cores = parallel::detectCores()) #<< 
rstan_options(auto_write = TRUE) #<<
```

Stan user's guide https://mc-stan.org/docs/2_18/stan-users-guide/index.html

---

## Installation of RStan

Follow instruction in https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started:

- Remove existing `RStan`, if any:

```{r, eval = F}
remove.packages("rstan")
if (file.exists(".RData")) file.remove(".RData")
```

- Restart `R` and install `rstan` package:

```{r, eval = F}
install.packages("rstan", repos = "https://cloud.r-project.org/", 
                 dependencies = TRUE)
```

- Make sure you have the necessary C++ toolchain: 

```{r}
pkgbuild::has_build_tools(debug = TRUE)
```

---

## Example: simple linear regression

`mtcars` is one of built-in data sets in `R`. 

```{r}
head(mtcars) %>% 
  rownames_to_column("name") %>%
  gt(rowname_col = vars(name))
```

---

## Example: simple linear regression

Scatter plot showing the relation between `wt` and `mpg`:

```{r, out.width = "50%"}
mtcars %>% ggplot(aes(wt, mpg)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

---

## Example: simple linear regression

Simple linear regression using `lm()`:

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
tidy(fit) %>% 
  gt(rowname_col = "term") %>%
  fmt_number(columns = vars(estimate, std.error, statistic), 
             decimals = 2) %>%
  fmt_scientific(columns = vars(p.value), decimals = 2)
```

---

## Example: simple linear regression

In particular, the 95% confidence intervals of the coefficients are given below:

```{r}
confint(fit) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  gt(rowname_col = "term") %>%
  fmt_number(columns = everything(), 
             decimals = 2)
```

---

## Example: simple linear regression

The following model is saved as `simple_lm.stan`:

```{bash, comment = ""}
cat simple_lm.stan
```

---

## Example: simple linear regression

Data need to be provided:

```{r stan, cache = T}
mtcarsData <- list(N = nrow(mtcars),
                   y = mtcars$mpg,
                   x = mtcars$wt)
fit_stan <- stan(file = 'simple_lm.stan', data = mtcarsData)
```

Recall that previously the following was obtained using `lm()`:

```{r, echo = F}
tidy(fit) %>% 
  gt(rowname_col = "term") %>%
  fmt_number(columns = vars(estimate, std.error, statistic), 
             decimals = 2) %>%
  fmt_scientific(columns = vars(p.value), decimals = 2)
```

```{r, echo = F}
confint(fit) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  gt(rowname_col = "term") %>%
  fmt_number(columns = everything(), 
             decimals = 2)
```

---

## Example: simple linear regression

```{r, comment = ""}
print(fit_stan)
```

---

## Example: simple linear regression

Run `stan()` with options:

```{r stan2, cache = T}
fit_stan2 <- stan(file = 'simple_lm.stan', 
                 data = mtcarsData,
                 seed = 2019,
                 chains = 5,
                 iter = 10000,
                 warmup = 3000)
```

---

## Example: simple linear regression

```{r, comment = ""}
print(fit_stan2)
```

---

## Example: simple linear regression

```{r, comment = "", out.width = "60%"}
traceplot(fit_stan2)
```

---

## Example: simple linear regression

Output from `stan()` can be extracted:

```{r}
df <- as.data.frame(fit_stan2)
dim(df) # iter = 10000, warmup = 3000, chains = 5
head(df)
```

---

## Example: logistic regression

```{r}
# virginica or not
dIris <- iris %>%
  mutate(IsVirginica = ifelse(Species == "virginica", 1, 0))

fit <- glm(IsVirginica ~ Sepal.Length + Sepal.Width, 
           data = dIris,
           family = binomial(link = "logit"))

tidy(fit) %>% 
  gt(rowname_col = "term") %>%
  fmt_number(columns = vars(estimate, std.error, statistic), 
             decimals = 2) %>%
  fmt_scientific(columns = vars(p.value), decimals = 2)
```

---

## Example: logistic regression

```{r}
confint(fit) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  gt(rowname_col = "term") %>%
  fmt_number(columns = everything(), 
             decimals = 2)
```

---

## Example: logistic regression

The following model is saved as `logistic_2.stan`:

```{bash, comment = ""}
cat logistic_2.stan
```

---

## Example: logistic regression

```{r logit, cache = T}
irisData <- list(N = nrow(dIris),
                 y = dIris$IsVirginica,
                 x1 = dIris$Sepal.Length,
                 x2 = dIris$Sepal.Width)
fit_logit <- stan(file = 'logistic_2.stan',
                  data = irisData,
                  seed = 2019,
                  chains = 5,
                  iter = 10000,
                  warmup = 3000)
```

---

## Example: logistic regression

```{r, comment = ""}
print(fit_logit)
```

---

## Example: logistic regression

```{r, comment = "", out.width = "60%"}
traceplot(fit_logit)
```
