---
title: "Lecture 1"
author: "Jung-Jin Lee"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "Stan.css"] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, message = F, echo = F, warning = F}
library(tidyverse) 
library(knitr) 
opts_chunk$set(
  tidy = FALSE,
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 3,
  fig.height = 3
  )
```

## Installation

Follow instruction in https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started:

- Remove existing `RStan`, if any:

```{r, eval = F}
remove.packages("rstan")
if (file.exists(".RData")) file.remove(".RData")
```

- Restart `R` and install `rstan` package:

```{r, eval = F}
install.packages("rstan", repos = "https://cloud.r-project.org/", 
                 dependencies = TRUE)
```

- Make sure you have the necessary C++ toolchain: 

```{r}
pkgbuild::has_build_tools(debug = TRUE)
```

---

## Example: simple linear regression

`mtcars` is one of built-in data sets in `R`. 

```{r}
head(mtcars)
dim(mtcars)
names(mtcars)
```

---

## Example: simple linear regression

Scatter plot showing the relation between `wt` and `mpg`:

```{r, out.width = "50%"}
mtcars %>% ggplot(aes(wt, mpg)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
  
```

---

## Example: simple linear regression

Simple linear regression using `lm()`:

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
summary(fit)$coef
```

In particular, the estimate for the coefficient of `wt` is `r fit$coef[2]`.

```{r}
confint(fit)
```


