---
title: "Homework 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

\bigskip

1. Compute
$$1^2+2^2+3^2+\cdots+99^2+100^2.$$

```{r, echo = T, eval = T}
vec <- 1:100
sum(vec^2)
```

\bigskip

2. How many digits does $N=1234567891011121314...9998999910000$ have, where $N$ is formed by concatenating $1,2,3,\ldots,9999,10000$? (_Hint_: you might want to look up the functions \texttt{nchar(), as.character(), paste()})

```{r, echo = T, eval = T}
vec_char <- as.character(1:10000) # turn numeric to character
vec_collapse <- paste(vec_char, collapse = "")
print(nchar(vec_collapse))
```

\bigskip

`iris` is a data set that comes with base R. It consists of `r nrow(iris)` observations with `r ncol(iris)` columns, `r names(iris)`, which are features related to Iris flowers. You can view the first few rows using `head()`:

```{r}
head(iris)
```

3. Compute the sample mean of `Sepal.Length`. Also, what is the _fifth_ smallest value of `Sepal.Length`? 

```{r, echo = T, eval = T}
mean(iris$Sepal.Length)
sort(iris$Sepal.Length)[5]
```

4. Obtain the following plots that describe the histogram of `Sepal.Length` by `Species`. 

```{r, echo = T, message=F, fig.height = 3}
g <- ggplot(iris, aes(Sepal.Length)) + 
  geom_histogram() +
  facet_wrap(~Species) 

print(g)
```

```{r, echo = FALSE}
d <- read.table(file = "heights.txt", header = T, sep = " ")
```

5. Recall that the data set `heights.txt` consists of `r nrow(d)` observations of mother/daughter height pairs. As done in class, you can read in the data set using the following code:

```{r}
d <- read.table(file = "heights.txt", header = T, sep = " ")
```

In how many pairs is mother's height greater than daughter's height?

```{r, echo = T, eval = T}
sum(d$Mheight > d$Dheight)
```


