---
title: Netlify blog workflow
author: Jung-Jin Lee
date: '2018-11-17'
categories:
  - Netlify
tags:
  - GitHub
slug: netlify-blog-workflow
---

R package blogdown과 Netlify를 이용한 블로그 개설/관리 과정을 여기에 기록해 둠

How to create a blog using blogdown and Netlify

1. Create a new repository: https://github.com/junglee0713/netlify 

2. Clone the repository, e.g.

```{bash, eval = F}
cd /Users/leej39/Dropbox/GitHub
git clone git@github.com:junglee0713/netlify.git 
```

3. (optional) Create an R project, say `/Users/leej39/Dropbox/GitHub/netlify/netlify.Rproj`.

4. (do only once) Open RStudio and install necessary packages, following instruction in https://bookdown.org/yihui/blogdown/installation.html

```{r, eval = F}
install.packages("blogdown")
blogdown::install_hugo()
```

5. Make sure that `/Users/leej39/Dropbox/GitHub/netlify` does not include files other than `LICENSE`, the RStudio project file (`*.Rproj`), `README` or `README.md`. If you are okay with the default theme, run the following 

```{r, eval = F}
blogdown::serve_site() # or blogdown::serve_site(theme = "user/repo") to choose a different theme
blogdown::hugo_build()
```

6. Change the configuration file `/Users/leej39/Dropbox/GitHub/netlify/config.toml` as appropriate.

7. If you want to delete example or unnecessary posts that came with the default theme, you can do so by deleting the corresponding `.md` or `.html` files under `/Users/leej39/Dropbox/GitHub/netlify/content`.

8. Stage changes in `/Users/leej39/Dropbox/GitHub/netlify`, commit, and push. 

9. Go to https://www.netlify.com/ and link the GitHub repository https://github.com/junglee0713/netlify so that Netlify can sync with GitHub and deploy any changes in the repository real time. 

10. To create a new post, click on **Addins** in RStudio and use **New Post**. Use **Update Metadata** to update the metadata associated with the post.

11. As noted in https://bookdown.org/yihui/blogdown/static-files.html, all files under the `static/` directory are copied to `public/` when Hugo renders a website. Therefore one needs to be cautious about files under `static/`. Also to build `.Rmd` documents with custom output formats, i.e., `.Rmd` documents not using the `blogdown::html_page()` format, a custom build script `R/build.R` must be provided with a single line of code in it:

```{r, eval = F}
blogdown::build_dir("static")
```



