---
title: Principal component analysis
author: Jung-Jin Lee
date: '2018-12-08'
slug: principal-component-analysis
---

```{r, echo = F}
knitr::opts_chunk$set(
  message = F
)
```

```{r, echo = F}
n <- 100
```

A mock data set consisting of `r n` data points, say $\mathbf{P}_1, \cdots, \mathbf{P}_{`r n`}$, is generated below. They depict a perturbed and slanted ellipse in $\mathbb{R}^3$. Let $X_{raw}$ be the $`r n` \times 3$ matrix whose rows consist of $\mathbf{P}_1, \cdots, \mathbf{P}_{`r n`}$. 

```{r}
set.seed(2018)
n <- 100
## Create a perturbed and slanted ellipse in R^3
x_center <- 3; y_center <- 1; z_center <- 2 
x_stretch <- 4; y_stretch <- 3; z_stretch <- 1
angle <- pi/3

x_ori <- x_stretch*cos(2*pi*(1:n)/n) + rnorm(n, mean = 0, sd = 0.5) # stretch
y_ori <- y_stretch*sin(2*pi*(1:n)/n) + rnorm(n, mean = 0, sd = 0.5) # stretch

x <- cos(angle)*x_ori - sin(angle)*y_ori + x_center # rotate
y <- sin(angle)*x_ori + cos(angle)*y_ori + y_center # rotate
z <- z_stretch*sin(2*pi*(1:n)/n) + z_center # make it slanted

Xraw <- data.frame(x, y, z) 
Xraw <- Xraw[sample(nrow(Xraw)), ] # permute rows to randomize P1, P2, P3
dim(Xraw)
```

The following is an interactive 3D plot showing the distribution of points:

```{r}
library(plotly)
colored <- c("P1", "P2", "P3", rep("P4 - P100", n-3))
plot_ly(Xraw, x = ~x, y = ~y, z = ~z, color = ~colored, 
        marker = list(size = 5), 
        colors = c("red", "blue", "green", "grey")) %>%
  add_markers()

```

Our goal is to find a _good_ (meaning to be clarified <a href="#later">later</a>) two dimensional representation $\mathbf{Q}_1, \cdots, \mathbf{Q}_{`r n`}$ of $\mathbf{P}_1, \cdots, \mathbf{P}_{`r n`}$ so that we can reduce the dimension (from 3 to 2) of data and visualize them. If $x$, $y$, and $z$ variables need to be equally weighted, then the data points should be centered and normalized. Let $X$ be the centered and normalized version of $X_{raw}$.

```{r}
X <- scale(Xraw) %>%
  as.data.frame()
plot_ly(X, x = ~x, y = ~y, z = ~z, color = ~colored, 
        marker = list(size = 5), 
        colors = c("red", "blue", "green", "grey")) %>%
  add_markers() 
```

<a name="later"></a> Consider a projection $P$ onto a two dimensional subspace $W$ of $\mathbb{R}^3$ and let $Y=XP$. Since $X$ is centered, so is $Y$ and hence the total variance of $Y$ is given by $\frac{1}{n-1}\|Y\|_{HS}^2$, where $\|\cdot\|_{HS}$ denotes the Hilbert-Schmidt norm. 
By a good two dimensional representation, we mean finding a projection $P$ such that $\frac{1}{n-1}\|Y\|_{HS}^2$ is maximized, where $Y=XP$. As $P$ is a projection matrix, $P=P^t=P^2$ and it follows that 
$$\|Y\|_{HS}^2=tr((XP)^t(XP))=tr(P^tX^tXP)=tr(X^tXPP^t)=tr(X^tY)$$ and that

\begin{aligned}
\|X-Y\|_{HS}^2&=tr((X-Y)^t(X-Y))\\
&=tr(X^tX)-tr(X^tY)-tr(Y^tX)+tr(Y^tY)\\
&=tr(X^tX)-tr(X^tY)-tr((Y^tX)^t)+tr(Y^tY)\\
&=tr(X^tX)-2tr(X^tY)+tr(Y^tY)\\
&=\|X\|_{HS}^2-\|Y\|_{HS}^2,
\end{aligned}

which implies that $\|Y\|_{HS}^2$ is maximized when $\|X-Y\|_{HS}^2$ is minimized. 

Let $X=U\Sigma V^t$ be a [singular value decomposition](https://en.wikipedia.org/wiki/Singular_value_decomposition), where $U$ is a $`r n` \times `r n`$ orthogonal matrix, $V$ is a $3 \times 3$ orthogonal matrix, and 
$$\Sigma=\left[\begin{array}{ccc} \sigma_1 & & \\ & \sigma_2 &  \\ & & \sigma_3 \\ & & \\ & & \end{array}\right]_{`r n`\times 3}$$ with
$$\sigma_1\geq \sigma_2\geq \sigma_3.$$
On one hand, by [Eckart–Young–Mirsky theorem](https://en.wikipedia.org/wiki/Low-rank_approximation), 
$$\min_{rank(Y)=2} \|X-Y\|_{HS}^2 = \sigma_3^2$$
On the other hand, let $Y_0=XP_0$, where $P_0=VQV^t$ and 
$$Q=\left[\begin{array}{ccc} 1 & & \\ & 1 &  \\ & & 0 \end{array}\right],$$
then $P_0$ is a projection of rank 2 satisfying $\|X-Y_0\|_{HS}^2 = \sigma_3^2$. This is our desired projection!

Note that $XP_0=U\Sigma V^tVQV^t = U\Sigma_0V^t$, where $$\Sigma_0=\left[\begin{array}{ccc} \sigma_1 & & \\ & \sigma_2 &  \\ & & 0 \\ & & \\ & & \end{array}\right]_{`r n` \times 3}.$$

In particular, this shows that the first two columns of $V$ are basis for $W$, the subspace associated with $P$, and the first two columns of $U\Sigma_0$ provide the desired coordinates $\mathbf{Q}_1, \cdots, \mathbf{Q}_{`r n`}$. This can be obtained as in the following using an R function `svd()`:

```{r}
SVD <- svd(X, nu = n)
U <- SVD$u # 100 x 100
Sigma <- rbind(diag(SVD$d), matrix(0, nrow = n-3, ncol = 3)) # 100 x 3
V <- SVD$v
X_recover <- (U %*% Sigma) %*% t(V) # This recovers X 
Q <- U %*% Sigma # The first two columns of Q gives coordinates Q1, Q2, ..., Q100
head(Q[, 1:2])
```

**Remark**: $U\Sigma$ (and hence $U\Sigma_0$) can be computed from the [orthogonal diagonalization](https://en.wikipedia.org/wiki/Diagonalizable_matrix#Diagonalizable_matrices) of $XX^t$, as  
$$XX^t = U\Sigma V^t(V\Sigma^tU^t) = U\Lambda U^t,$$
where $$\Lambda = \Sigma\Sigma^t= \left[\begin{array}{ccccc} \sigma_1^2 & & & &\\ & \sigma_2^2 & & & \\ & & \sigma_3^2 & & \\ & & & 0 & \\ & & & & \ddots  \end{array}\right]_{`r n` \times `r n`}.$$

The R function `prcomp()` easily computes the coordinates $\mathbf{Q}_1, \cdots, \mathbf{Q}_{`r n`}$ from $X_{raw}$:

```{r}
Y <- prcomp(Xraw, center = T, scale = T)$x %>%
  as.data.frame() # PC1 and PC2 of Y represent coordinates of Q1, Q2, ..., Q100 
head(Y)
```

The projected points on the PC1-PC2 plane viewed from PC3 axis:

```{r}
scene <- list(camera = list(up = list(x = 1, y = 0, z = 0),
                           eye = list(x = 0, y = 0, z = 2)))
plot_ly(Y, x = ~PC1, y = ~PC2, z = ~PC3, color = ~colored,
        marker = list(size = 5), 
        colors = c("red", "blue", "green", "grey")) %>%
  add_markers() %>%
  layout(scene = scene)
```

Finally, one can get the following 2D plot as a result of dimension reduction: 

```{r}
Y %>% mutate(color_code = colored) %>%
  ggplot(aes(PC1, PC2, color = color_code)) +
  scale_color_manual(values = c("red", "blue", "green", "grey")) +
  theme(aspect.ratio = 1) +
  geom_point()
```

