---
title: anova( ) and Anova( )
author: Jung-Jin Lee
date: '2019-03-17'
slug: anova-and-anova
categories: []
tags: []
---

```{r, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  message = F,
  warning = F
)
```

After spending hours of despair and confusion, I came across an [excellent illustration given at Stack Exchange](https://math.stackexchange.com/questions/1792351/sequential-anova-r).

There are various $p$-values coming out of `lm()`, `anova()`, and `Anova()`. First of all, `car` library is loaded for use of `Anova()`:

```{r}
library(car) # for Anova()
```

The built-in `mtcars` will be used as an example.

```{r}
attach(mtcars)
head(mtcars)
```

Various linear models for comparison, the full model being `mpg ~ cyl + hp + wt`.

```{r}
fit_null <- lm(mpg ~ 1) 
fit_cyl <- lm(mpg ~ cyl)
fit_hp <- lm(mpg ~ hp)
fit_wt <- lm(mpg ~ wt)
fit_cyl_hp <- lm(mpg ~ cyl + hp)
fit_cyl_wt <- lm(mpg ~ cyl + wt)
fit_hp_wt <- lm(mpg ~ hp + wt)
fit_full <- lm(mpg ~ cyl + hp + wt)
```

We are starting off with the summary of the full model:

```{r}
summary(fit_full)
```

First of all, the $p$-value `2.184e-11` in the bottom comes from comparing the null model to the full model:

```{r}
anova(fit_null, fit_full)
```

The $p$-values from the coefficients are from comparing the full model with the model missing the respective term:

```{r}
anova(fit_hp_wt, fit_full) # missing cyl
anova(fit_cyl_wt, fit_full) # missing hp
anova(fit_cyl_hp, fit_full) # missing wt
```

These $p$-values are the same as what is obtained from `Anova()`:

```{r}
Anova(fit_full)
```

However, notice that `anova()` <font color="red"> **does not** </font> give the same result:

```{r}
anova(fit_full)
```

Well, at least this result is _sequential_ as can be seen in the following:

```{r}
anova(fit_null, fit_cyl, fit_cyl_hp, fit_full)
```

Where did the $p$-values `5.093e-12`, `0.1185183`, and `0.0001995` come from? Maybe comparing `fit_null` vs `fit_cyl`, `fit_cyl` vs `fit_cyl_hp`, and `fit_cyl_hp` vs `fit_full`, respectively?  

```{r}
anova(fit_null, fit_cyl)
anova(fit_cyl, fit_cyl_hp)
anova(fit_cyl_hp, fit_full)
```

Only the last one gives the same $p$-value. Why is it?

This is because `anova()` uses a different $F$ statistic to compute $p$-value when a `lm` object with two or more predictors is provided as a single argument -- again, refer to an [excellent illustration given at Stack Exchange](https://math.stackexchange.com/questions/1792351/sequential-anova-r).

Note that `anova()` and `Anova()` agree in case of simple linear regression:

```{r}
summary(fit_cyl)
anova(fit_cyl)
Anova(fit_cyl)
```



