---
title: Distance between groups
author: Jung-Jin Lee
date: '2019-10-25'
slug: distance-between-groups
categories: []
tags: []
---

```{r, message = FALSE}
library(usedist)
library(tidyverse)
library(ggforce)
library(ggsci)
library(gt)
library(ggrepel)
library(ggdendro)
```

First, we make a fake data set:

```{r}
set.seed(2019)
n <- 6
ADf <- tibble(x = rnorm(n, sample(10, 1), sample(2, 1)),
              y = rnorm(n, sample(10, 1), sample(2, 1)), 
              group = "A")
BDf <- tibble(x = rnorm(n, sample(10, 1), sample(2, 1)), 
              y = rnorm(n, sample(10, 1), sample(2, 1)), 
              group = "B")
CDf <- tibble(x = rnorm(n, sample(10, 1), sample(2, 1)), 
              y = rnorm(n, sample(10, 1), sample(2, 1)), 
              group = "C")
DDf <- tibble(x = rnorm(n, sample(10, 1), sample(2, 1)), 
              y = rnorm(n, sample(10, 1), sample(2, 1)), 
              group = "D")
EDf <- tibble(x = rnorm(n, sample(10, 1), sample(2, 1)), 
              y = rnorm(n, sample(10, 1), sample(2, 1)), 
              group = "E")
df <- bind_rows(ADf, BDf, CDf, DDf, EDf) %>%
  group_by(group) %>%
  mutate(SampleID = paste0(group, 1:n())) %>%
  ungroup()
```

Visualization of the data:

```{r}
ggplot(df, aes(x, y)) + 
  geom_point(aes(shape = group), show.legend = FALSE) +
  geom_mark_hull(aes(label = group, fill = group), 
                 show.legend = FALSE) +
  scale_fill_npg() +
  theme_void() 
```

Make a `distance` object:

```{r}
Dist <- dist(df %>% 
               select(-group) %>% 
               column_to_rownames("SampleID")) ## euclidean
```

Which samples are in which group?

```{r}
Gr <- df$group
names(Gr) <- df$SampleID
```

Make sure the order of the grouping is the same as `Dist`:

```{r}
Gr <- Gr[colnames(as.matrix(Dist))]
```

Create a data frame of distances between groups of items using `dist_groups` from `usedist` package (https://github.com/kylebittinger/usedist):

```{r}
Dist_Gr <- dist_groups(Dist, Gr)
Dist_Gr %>%
  head() %>%
  gt()
```

What are the distances between groups?

```{r}
Dist_Between_Groups <- Dist_Gr %>%
  arrange(Label, Distance) %>%
  filter(!duplicated(Label)) %>%
  mutate(Distance = ifelse(str_detect(Label, "^Within "), 0, Distance)) %>%
  mutate(rowItem = str_remove_all(Label, "^Between | and .+|^Within ")) %>%
  mutate(colItem = str_remove_all(Label, "^.+ and |^Within ")) %>%
  select(Item1, Item2, rowItem, colItem, Label, Distance) 

Dist_Matix <- Dist_Between_Groups %>%
  select(-Item1, -Item2, -Label) %>%
  pivot_wider(names_from = "colItem", values_from = "Distance", values_fill = list(Distance = 0)) %>%
  column_to_rownames("rowItem")

Dist_Matix <- Dist_Matix[, rownames(Dist_Matix)] # arrange columns
Dist_Matix <- Dist_Matix + t(Dist_Matix) # make a symmetric matrix
```

Now distances between groups are:

```{r}
Dist_Matix %>%
  rownames_to_column() %>%
  gt(rowname_col = "rowname")
```

Which samples achieve the minimum distances? How about dendrogram?

```{r}
whichSamplesFirst <- Dist_Between_Groups %>%
  select(Item1, Label) %>%
  mutate(Item1 = as.character(Item1)) %>%
  left_join(df %>% select(-group), by = c("Item1" = "SampleID")) %>%
  rename(x1 = x, y1 = y)

whichSamplesSecond <- Dist_Between_Groups %>%
  select(Item2, Label) %>%
  mutate(Item2 = as.character(Item2)) %>%
  left_join(df %>% select(-group), by = c("Item2" = "SampleID")) %>%
  rename(x2 = x, y2 = y)

whichSampleCombined <- whichSamplesFirst %>%
  left_join(whichSamplesSecond, by = "Label") %>%
  filter(str_detect(Label, "^Between "))

g_distance <- df %>% 
  mutate(repel_label = ifelse(SampleID %in% c(whichSampleCombined$Item1, whichSampleCombined$Item2),
                              SampleID, "")) %>%
  ggplot(aes(x, y)) + 
  geom_point(aes(shape = group), show.legend = FALSE) +
  geom_mark_hull(aes(label = group, fill = group), 
                 show.legend = FALSE) +
  scale_color_d3() +
  scale_fill_npg() +
  geom_text_repel(aes(label = repel_label)) +
  geom_segment(data = whichSampleCombined, 
               aes(x = x1, y = y1, xend = x2, yend = y2, color = Label)) +
  theme_void() 

hc <- hclust(as.dist(Dist_Matix))
hcdata <- dendro_data(hc, type = "rectangle")

g_dendro <- segment(hcdata) %>%
  ggplot() + 
  geom_segment(aes(x, y, xend = xend, yend = yend)) +
  geom_text(data = hcdata$labels, aes(x, y, label = label),
            hjust = -1, size = 3) +
  coord_flip() + 
  scale_y_reverse() +
  theme_classic() +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title = element_blank())
```

```{r}
g_distance
```

```{r, fig.height = 2, fig.width = 3.8, fig.align = "center"}
g_dendro
```

