---
title: "Lecture 4"
author: "Jung-Jin Lee"
date: "Jan 29, 2019"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "Drexel.css"] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
 
```{r, message = F, echo = F, warning = F}
library(tidyverse) 
library(gridExtra)
library(knitr) 
opts_chunk$set(
  tidy = FALSE,
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 4,
  fig.height = 4
  )
```

## Normal distribution

Consider the data set `heights.txt` which consists of mother/daughter height pairs. 

```{r}
ht <- read.table(file = "heights.txt", header = T, sep = " ")
dim(ht)
head(ht)
```

---

## Normal distribution, continued

.pull-left[
```{r}
ggplot(ht, aes(Mheight)) + 
  geom_histogram()
```
]

.pull-right[
```{r}
ggplot(ht, aes(Dheight)) + 
  geom_histogram()
```
]

Note that the histograms look like a bell.

---

## Normal distribution, continued

Bell shapes are observed in other data sets, too. 

.pull-left[
```{r}
ggplot(diamonds, aes(depth)) + 
  geom_histogram()
```
]

--

.pull-right[
```{r}
ggplot(iris, aes(Sepal.Width)) + 
  geom_histogram() 
```
]

---

## Rigorous definition of normal distribution

- A continuous random variable $X$ is said to follow a _normal distribution_ $N(\mu, \sigma^2)$ if the probability $P(a\leq X \leq b)$ that $X$ is between $a$ and $b$ is given by 
$$P(a\leq X \leq b) = \int_a^b \phi_{\mu,\sigma}(x)\,dx,$$
where $$\phi_{\mu,\sigma}(x) = \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{(x-\mu)^2}{2\sigma^2}}.$$

--

- In particular, the area under the curve equals 1.

--

- $\phi_{\mu,\sigma}(x)$ is a bell-shaped function called the probability distribution function (pdf) of $X$. $\mu$ is called the _mean_ and $\sigma$ is called the _standard deviation_.

--

- In particular, if $\mu = 0$ and $\sigma = 1$, $X$ is said to follow the _standard_ normal distribution.

---

## Why is normal distribution important?

- Has mathematically important properties (e.g. Central Limit Theorem).

- Related with other important distributions (e.g. $t$, $F$, $\chi^2$).

- Provides a good model for many real-world data (e.g. A normal distribution is a good model for data such as heights or weights).

---

## Normal pdf

```{r, echo = FALSE}
fun.1 <- function(x) dnorm(x, mean = 3, sd = 2)
g1 <- ggplot(data.frame(x = 3), aes(x)) +
  scale_x_continuous(lim = c(-3, 9), breaks = c(-3, 3, 9)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.1) +
  ylim(c(0, 0.4)) +
  theme(aspect.ratio = 1) +
  geom_segment(x = 3, y = 0, xend = 3, yend = 0.2, lty = "dashed", color = "red") +
  ggtitle(bquote('N(3,'~2^2~')'))

fun.2 <- function(x) dnorm(x, mean = 0, sd = 1)
g2 <- ggplot(data.frame(x = 0), aes(x)) +
  scale_x_continuous(lim = c(-2.5, 2.5), breaks = c(-2, 0, 2)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.2) + 
  ylim(0, 0.4) +
  theme(aspect.ratio = 1) +
  geom_segment(x = 0, y = 0, xend = 0, yend = 0.4, lty = "dashed", color = "red") +
  ggtitle(bquote('N(0,'~1^2~')'))
```

.pull-left[
```{r, echo = F}
print(g1)
```
]

--

.pull-right[
```{r, echo = F}
print(g2)
```
]

_mean_ determines the center of the bell shape, while _standard deviation_ measures the spread.

---

## Normal pdf, continued

**Example**: the life of a fully-charged cell phone battery is normally distributed with a mean of 13 hours with a standard deviation of 1 hour. $X$, the life of a randomly selected fully-charged phone battery, can be assumed to follow $N(13, 1^2)$. 

```{r, echo = F, out.width = "50%"}
fun.3 <- function(x) dnorm(x, mean = 13, sd = 1)
g3 <- ggplot(data.frame(x = 13), aes(x)) +
  scale_x_continuous(lim = c(9, 17), breaks = c(10, 13, 16)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.3) + 
  ylim(0, 0.4) +
  theme(aspect.ratio = 1) +
  geom_segment(x = 0, y = 0, xend = 0, yend = 0.4, lty = "dashed", color = "red") +
  ggtitle(bquote('N(13,'~1^2~')'))
print(g3)
```

---

## Normal pdf, continued

- What is the probability that $X \leq 13$?

--

```{r, echo = F, out.width = "50%"}
g4 <-ggplot(data.frame(x = 13), aes(x)) +
  scale_x_continuous(lim = c(9, 17), breaks = c(10, 13, 16)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.3, xlim = c(9, 13), geom = "area", fill = "red") + 
  stat_function(fun = fun.3) + 
  ylim(0, 0.45) +
  theme(aspect.ratio = 1) +
  geom_segment(x = 0, y = 0, xend = 0, yend = 0.4, lty = "dashed", color = "red") +
  ggtitle(bquote('N(13,'~1^2~')'))
print(g4)
```

--

- As the curve is symmetric, the desired probability is $0.5$.

---

## Normal pdf, continued

- What is the probability that $X \leq 12$?

.pull-left[
```{r, echo = F}
g5 <- ggplot(data.frame(x = 12), aes(x)) +
  scale_x_continuous(lim = c(9, 17), breaks = c(12, 13)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.3, xlim = c(9, 12), geom = "area", fill = "red") + 
  stat_function(fun = fun.3) + 
  ylim(0, 0.45) +
  theme(aspect.ratio = 1) +
  geom_segment(x = 0, y = 0, xend = 0, yend = 0.4, lty = "dashed", color = "red") +
  ggtitle(bquote('N(13,'~1^2~')'))
print(g5)
```
]

--

.pull-right[
`pnorm()` computes the desired probability:
```{r}
pnorm(12, mean = 13, sd = 1)
pnorm(13, mean = 13, sd = 1)
```
]

--

**Exercise**: compute the probability that X $\geq$ 14.

---

## Normal pdf, continued

- What is the probability that $11 \leq X \leq 14$?

.pull-left[
```{r, echo = F}
g6 <- ggplot(data.frame(x = 13), aes(x)) +
  scale_x_continuous(lim = c(9, 17), breaks = c(11, 14)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.3, xlim = c(11, 14), geom = "area", fill = "purple") + 
  stat_function(fun = fun.3) + 
  ylim(0, 0.45) +
  theme(aspect.ratio = 1) +
  ggtitle(bquote('N(13,'~1^2~')'))
print(g6)
```
]

--

.pull-right[
```{r, echo = F}
g7 <- ggplot(data.frame(x = 13), aes(x)) +
  scale_x_continuous(lim = c(9, 17), breaks = c(11, 14)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.3, xlim = c(11, 14), geom = "area", fill = "red", alpha = 0.5) + 
  stat_function(fun = fun.3, xlim = c(9, 14), geom = "area", fill = "blue", alpha = 0.5) +
  stat_function(fun = fun.3) + 
  ylim(0, 0.45) +
  theme(aspect.ratio = 1) +
  ggtitle(bquote('N(13,'~1^2~')'))
print(g7)
```
]

---

## Normal pdf, continued

**Exercise**: 

- Compute the probability that $X\geq 15$, where $X$ follows $N(13,1^2)$.
- Compute the probability that $Z \leq -1$, where $Z$ follows $N(0,1^2)$.
- Determine $z_{0.1}$ such that the probability that $Z\leq z_{0.1}$ equals 0.9 (_hint_: look up `qnorm()`). 

.pull-left[
```{r, echo = F, out.width = "90%"}
fun.4 = function(x) dnorm(x, mean = 0, sd = 1)
g8 <- ggplot(data.frame(x = 0), aes(x)) +
  scale_x_continuous(lim = c(-2, 2), breaks = c(-1, 0)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.4, xlim = c(-2,-1), geom = "area", fill = "red") + 
  stat_function(fun = fun.4) + 
  ylim(0, 0.45) +
  theme(aspect.ratio = 1) +
  ggtitle(bquote('N(0,'~1^2~')'))
print(g8)
```
]

.pull-right[
```{r, echo = F, out.width = "90%"}
g9 <- ggplot(data.frame(x = 0), aes(x)) +
  scale_x_continuous(lim = c(-2, 2), breaks = c(1.28), labels = c(bquote('z'[0.1]))) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = fun.4, xlim = c(-2, 1.28), geom = "area", fill = "red") + 
  stat_function(fun = fun.4) + 
  ylim(0, 0.45) +
  theme(aspect.ratio = 1) +
  ggtitle(bquote('N(0,'~1^2~')'))
print(g9)
```
]

---

## _t_-distribution

- Class of bell-shaped distribution determined by a parameter called _degrees of freedom_.

```{r, echo = F, out.width = "90%", fig.height = 4, fig.width = 6}
tfun.1 = function(x) dt(x, df = 1)
tfun.5 = function(x) dt(x, df = 5)
tfun.10 = function(x) dt(x, df = 10)
zfun = function(x) dnorm(x)

gt <- ggplot(data.frame(x = 0), aes(x)) +
  scale_x_continuous(lim = c(-2, 2), breaks = c(-2, 0, 2)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = tfun.1, aes(color = "a")) + 
  stat_function(fun = tfun.5, aes(color = "b")) + 
  stat_function(fun = tfun.10, aes(color = "c")) + 
  stat_function(fun = zfun, aes(color = "d")) + 
  ylim(0, 0.45) +
  scale_color_manual(name = "Functions",
                       values = c("blue", "red", "green", "black"), 
                       labels = c("t, df = 1", "t, df = 5", "t, df = 10", "Standard normal"))
  
print(gt)
```

---

## _t_-distribution, continued

**Example**: suppose $X$ follows $t$-distribution with $5$ degrees of freedom. What is the probability that $X\leq 0$? $X\leq -1.5$? $-1 \leq X \leq 1.2$? $X \geq 1.6$?

.pull-left[
```{r, echo = F}
gt5 <- ggplot(data.frame(x = 0), aes(x)) +
  scale_x_continuous(lim = c(-2, 2), breaks = c(-1.5, -1, 0, 1.2, 1.6)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = tfun.5) + 
  ylim(0, 0.45) +
  ggtitle("t, df = 5")
print(gt5) 
```
]

--

.pull-right[
```{r}
pt(0, df = 5)
pt(-1.5, df = 5)
pt(1.2, df = 5) - pt(-1, df = 5)
1-pt(1.6, df = 5)
```
]

---

## _t_-distribution, continued

**Exercise**: suppose $X$ follows $t$-distribution with $5$ degrees of freedom. Determine $t_{5, 0.2}$ so that the probability that $X\geq t_{5, 0.2} = 0.2$ (_Hint_: look up `qt()`).

```{r, echo = F, out.width = "55%"}
library(latex2exp)
gt5 <- ggplot(data.frame(x = 0), aes(x)) +
  scale_x_continuous(lim = c(-2, 2), breaks = c(0.92), labels = c(TeX("$t_{5, 0.2}$"))) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = tfun.5, xlim = c(0.92, 2), geom = "area", fill = "red") + 
  stat_function(fun = tfun.5) + 
  ylim(0, 0.45) +
  ggtitle("t, df = 5")
print(gt5) 
```

---

## _t_-distribution, continued

**Exercise**: suppose $X$ follows $t$-distribution with $4$ degrees of freedom. Determine $t_{4, 0.025}$ so that the probability that $-t_{4, 0.025} \leq X \leq t_{4, 0.025} = 0.95$.

```{r, echo = F, out.width = "55%"}
library(latex2exp)
tfun.4 = function(x) dt(x, df = 4)
gt4 <- ggplot(data.frame(x = 0), aes(x)) +
  scale_x_continuous(lim = c(-3.5, 3.5), breaks = c(-2.77, 2.77), labels = c(TeX("$-t_{4, 0.025}$"), TeX("$t_{4, 0.025}$"))) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = tfun.4, xlim = c(-2.77, 2.77), geom = "area", fill = "red") + 
  stat_function(fun = tfun.4) + 
  ylim(0, 0.45) +
  ggtitle("t, df = 4")
print(gt4) 
```

---

## Using data to make inference

- Point estimate

--

- Confidence interval

--

- Statistical tests

---

## Point estimate

- Assume that the heights of all UK women follow a normal distribution $N(\mu, \sigma^2)$.
- Interested in determining $\mu$ and $\sigma$.
- Sample mean ans sample standard deviation are known to be _good_ estimates for $\mu$ and $\sigma$, respectively. 

--

.pull-left[
```{r, echo = F, out.width = "95%"}
ggplot(ht, aes(Mheight)) + 
  geom_histogram()
```
]

--

.pull-right[
```{r}
mean(ht$Mheight)
sd(ht$Mheight)
```

**Exercise**: compute the sample mean and sample standard deviation of `Sepal.Width` in `iris` data set.

]

---

## Confidence interval

Point estimate is too naive -- how much can we trust `r mean(ht$Mheight)` as a reliable estimator for $\mu$, the true population mean?

**Example**: suppose that we are measuring the battery life which is assume to follow a normal distribution. Assume that we collected two sets of samples of size 5:
$$\text{Sample 1: }\{12, 13, 13, 14, 13\}$$
$$\text{Sample 2: }\{10, 11, 16, 13, 15\}$$
```{r}
sample1 <- c(12, 13, 13, 14, 13); sample2 <- c(10, 11, 16, 13, 15)
c(mean(sample1), mean(sample2))
c(sd(sample1), sd(sample2))
```

---

## Confidence interval, continued

Suppose that a sample of size $n$, say $X_1, X_2, \ldots, X_n$, is taken from a population that follows $N(\mu, \sigma^2)$. The 100 $\times (1-\alpha)$ % confidence interval of $\mu$ is defined as
$$\left(\bar{X} - t_{n-1, \alpha/2}\frac{S}{\sqrt{n}},~\bar{X} + t_{n-1, \alpha/2}\frac{S}{\sqrt{n}}\right),$$
where, $\bar{X}$ and $S$ represent the sample mean and sample standard deviation, respectively and $t_{n-1, \alpha/2}$ is the quantile value of a $t$-distribution of $n-1$ degrees of freedom with tail probability $\alpha/2$.

In particular, the 95% confidence interval (so $\alpha = 0.05$) of $\mu$ is given by 
$$\left(\bar{X} - t_{n-1, 0.025}\frac{S}{\sqrt{n}},~\bar{X} + t_{n-1, 0.025}\frac{S}{\sqrt{n}}\right),$$

Interpretation of 95% confidence interval: when this procedure of creating 95% confidence intervals is repeated, about 95% of the intervals will contain the true population mean. 

---

## Confidence interval, continued

```{r}
alpha <- 0.05
xbar <- mean(sample1)
sdev <- sd(sample1)
n <- length(sample1)
tval <- qt(1-alpha/2, df = n-1)
CI95 <- xbar + tval*sdev/sqrt(n)*c(-1,1)
CI95
```

**Exercise**: compute the 95% confidence interval of `sample2`. 

---

## Confidence interval, continued

`t.test()` can be used to compute confidence interval of mean.

```{r}
t1 <- t.test(sample1)
t1
```

---

## Confidence interval, continued

```{r}
t2 <- t.test(sample2)
t2
```

---

## Confidence interval, revisited

Review of confidence interval: 

$$\left(\bar{X} - t_{n-1, \alpha/2}\frac{S}{\sqrt{n}},~\bar{X} + t_{n-1, \alpha/2}\frac{S}{\sqrt{n}}\right)$$

- As $1-\alpha$ (called confidence level) increases, the confidence interval becomes wider. 

- As $n$ (sample size) increases, the confidence interval becomes narrower. 

---

## Exercises

**Exercise**: based on the `heights.txt` data, find the 80%, 90% and 95% confidence interval of the mean of UK women's heights (mom's heights).

**Exercise**: `starwars`, which comes with the package `dplyr`, is a data set that contains data related with characters in the movie Star Wars. 

- Explorer the data set. Subset the data so that the resulting data set contains only human species with `height` data available (_Hint_: look up `is.na()`). How many observations?

---

## Exercises, continued

- Explore the data set and create the following plot showing the height distribution of Human species:
```{r, echo = F, out.width = "50%"}
dd <- starwars %>% filter(species == "Human") %>%
  filter(!is.na(height)) 

ggplot(dd, aes(x = gender, y = height)) +
  geom_boxplot()
```

- Find the name of the tallest Human female character.

```{r, echo = F}
ddd <- dd %>% filter(gender == "female") %>%
  arrange(desc(height))
```

- Find the 95% confidence interval of the mean of Human male heights.
