---
title: "Lecture 7"
author: "Jung-Jin Lee"
date: "Feb 26, 2019"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "Drexel.css"] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
 
```{r, message = F, echo = F, warning = F}
library(tidyverse) 
library(gridExtra)
library(knitr) 
opts_chunk$set(
  tidy = FALSE,
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 4,
  fig.height = 4
  )
```

## _t_-test: review

- One-sample $t$-test

  - Given sample, test $H_0: \mu = \mu_0$.

- Two-sample $t$-test:

  - With equal variance assumption
  
  - Without equal variance assumption (Welch's test)
  
  - Test $H_0$: two populations have the same mean. 
  
- Paired $t$-test

  - Need paired samples
  
  - Test $H_0$: two populations have the same mean. 

- In all cases, use `t.test()`

---

## _F_-test

- Comparison of two variances (Levene's test)

- Comparison of two or more means (one-way ANOVA)

---

## _F_-distribution

```{r, echo = F, out.width = "90%", fig.height = 4, fig.width = 6}
library(latex2exp)
f_1_2 = function(x) df(x, df1 = 1, df = 2)
f_2_1 = function(x) df(x, df1 = 2, df = 1)
f_5_2 = function(x) df(x, df1 = 5, df = 2)
f_30_30 = function(x) df(x, df1 = 30, df = 30)
gf <- ggplot(data.frame(x = 1), aes(x)) +
  scale_x_continuous(lim = c(0, 5), breaks = 0:5) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = f_1_2, aes(color = "a")) + 
  stat_function(fun = f_2_1, aes(color = "b")) + 
  stat_function(fun = f_5_2, aes(color = "c")) + 
  stat_function(fun = f_30_30, aes(color = "d")) + 
  ylim(0, 1.2) +
  scale_color_manual(name = "Functions",
                       values = c("blue", "red", "green", "black"), 
                       labels = c("df1=1, df2=2", "df1=2, df2=1", "df1=5, df2=2", "df1=30, df2=30"))
print(gf) 
```

---

## Levene's test

**Example**: A biologist is interested in comparing the lengths of female and male green lynx spiders. Assume that the length of male spiders follows $N(\mu_m, \sigma_m^2)$ and the length female spiders follows $N(\mu_f, \sigma_f^2)$. 

```{r}
male = c(5.20, 4.70, 5.75, 7.50, 6.45, 6.55, 4.70, 4.80, 5.95, 
         5.20, 6.35, 6.95, 5.70, 6.20, 5.40, 6.20, 5.85, 6.80, 
         5.65, 5.50, 5.65, 5.85, 5.75, 6.35, 5.75, 5.95, 5.90, 
         7.00, 6.10, 5.80)
female = c(8.25, 9.95, 5.90, 7.05, 8.45, 7.55, 9.80, 10.80, 6.60, 
           7.55, 8.10, 9.10, 6.10, 9.30, 8.75, 7.00, 7.80, 8.00, 
           9.00, 6.30, 8.35, 8.70, 8.00, 7.50, 9.50)
c(length(male), length(female))
c(mean(male), mean(female))
```

---

## Levene's test, continued

```{r}
# ggplot needs a dataframe
male_df <- data.frame(sex = "male", length = male)
female_df <- data.frame(sex = "female", length = female)
spider <- rbind(male_df, female_df) #<<
```

```{r, echo = F, out.width = "50%"}
ggplot(spider, aes(sex, length)) + geom_boxplot()
```

---

## Levene's test, continued

What can we say about variances? 

.pull-left[
```{r}
v_df <- spider %>%
  group_by(sex) %>% #<<
  summarize(va = var(length)) #<<
v_df
```
```{r, eval = F}
ggplot(v_df, aes(sex, va)) +
  geom_col() #<<
```
]

.pull-right[
```{r, echo = F}
ggplot(v_df, aes(sex, va)) +
  geom_col() #<<
```
]

Can we say that the population variances are the same between male and female spiders? (i.e., $\sigma_m^2 = \sigma_f^2$?)

---

## Levene's test, continued

Suppose that $X_1,X_2,\ldots,X_n$ (respectively, $Y_1,Y_2,\ldots,Y_m$) are taken from a population that follows $N(\mu_X,\sigma_X^2)$ (respectively, $N(\mu_Y,\sigma_Y^2)$). Let 
$$\bar{X}=\frac{X_1+X_2+\cdots+X_n}{n}, \quad \bar{Y}=\frac{Y_1+Y_2+\cdots+Y_m}{m},$$ 
and
$$S_X^2=\frac{1}{n-1}\sum_{i=1}^n(X_i-\bar{X})^2, \quad S_Y^2=\frac{1}{m-1}\sum_{j=1}^m(Y_j-\bar{Y})^2.$$ 
Then
$$\frac{S_X^2/\sigma_X^2}{S_Y^2/\sigma_Y^2}$$ follows an $F$-distribution with $df1=n-1$ and $df2=m-1$ degrees of freedom.

---

## Levene's test, continued

Going back to `spider` data, under the hypothesis $H_0: \sigma_m^2 = \sigma_f^2$, the fraction
$$\frac{S_m^2/\sigma_m^2}{S_f^2/\sigma_f^2} = \frac{S_m^2}{S_f^2}$$ should follow an $F$-distribution with $$df1=30-1=29 \quad \text{and}\quad df2=25-1=24$$ degrees of freedom.

```{r}
n <- length(male); m <- length(female)
var_m <- var(male); var_f <- var(female)
f <- var_m/var_f
print(f)
```

How likely or unlikely is this?

---

## Levene's test, continued

Under the null hypothesis $H_0: \sigma_m^2 = \sigma_f^2$, one would expect that `f` is close to 1. 

```{r, echo = F, out.width = "90%", fig.height = 4, fig.width = 6}
f_29_24 = function(x) df(x, df1 = 12, df = 12)
gf <- ggplot(data.frame(x = 1), aes(x)) +
  scale_x_continuous(lim = c(0, 3.5), breaks = c(0, f, 1, 2)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = f_29_24) + 
  geom_segment(x = f, xend = f, y = 0, yend = f_29_24(f), 
               lty = "dotted", color = "red") +
  ylim(0, 1) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0.5)) +
  ggtitle(TeX("F_{df1=29, df2=24}"))
print(gf) 
```

---

## Levene's test, continued

The probability of an event that is as extreme as or more extreme than the current observation (red area) seems to be small.

```{r, echo = F, out.width = "90%", fig.height = 4, fig.width = 6}
f_29_24 = function(x) df(x, df1 = 12, df = 12)
gf <- ggplot(data.frame(x = 1), aes(x)) +
  scale_x_continuous(lim = c(0, 3.5), breaks = c(0, f, 1, 2)) +
  theme_bw() + 
  theme_classic() +
  stat_function(fun = f_29_24, xlim = c(0, f), geom = "area", fill = "red") + 
  stat_function(fun = f_29_24, xlim = c(3.1, 3.5), geom = "area", fill = "red") + 
  stat_function(fun = f_29_24) + 
  ylim(0, 1) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0.5)) +
  ggtitle(TeX("F_{df1=29, df2=24}"))
print(gf) 
```

---

## Levene's test, continued

```{r}
lower_red = pf(f, df1 = n-1, df2 = m-1)
pval = 2*lower_red
print(pval)
```

--

$p$-value can be easily obtained by `var.test()`:

```{r}
var.test(male, female)
```

---

## Aside: one-sample variance test

One-sample variance test is based on $\chi^2$-test and can be done using `VarTest()` in `DescTools` pacakge. 

```{r}
#install.packages("DescTools")
library(DescTools)
# To test if population variance equals 0.5
VarTest(male, sigma.squared = 0.5) 
```

---

## One-way ANOVA (ANalysis Of VAriance)

**Example**: Dogs were fed a dry diet coated with different agents that were believed to affect the build-up of calculus on the teeth. Calculus accumulation was measured by an index that combined estimates of both the proportion of the teeth covered by the deposit and the thickness of the deposit. 26 dogs were randomly allocated to three treatments: `control`, soluble pyrophosphate (`p2o7`), and sodium hexametaphosphate (`hmp`). The data are presented in the following:

```{r}
control <- c(0.49, 1.05, 0.79, 1.35, 0.55, 1.36, 1.55, 1.66, 1.00)
p2o7 <- c(0.34, 0.76, 0.45, 0.69, 0.87, 0.94, 0.22, 1.07, 1.38)
hmp <- c(0.34, 0.05, 0.53, 0.19, 0.28, 0.45, 0.71, 0.95)
```

Can we say that all the three means are equal?

---

## Box plots

```{r}
ctrl_df <- data.frame(diet = "Control", index = control)
pyro_df <- data.frame(diet = "Pyrophosphate", index = p2o7)
sod_hexa_df <- data.frame(diet = "Hexametaphosphate", index = hmp)
dogs <- rbind(ctrl_df, pyro_df, sod_hexa_df)
mean_df <- dogs %>%
  group_by(diet) %>%
  summarize(count = n(), mean_index = mean(index))
mean_df
```

