---
title: "Homework 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

\bigskip

1. Compute
$$1^2+2^2+3^2+\cdots+99^2+100^2$$

```{r, echo = F, eval = F}
vec <- 1:100
sum(vec^2)
```

\bigskip

2. How many digits does $N=1234567891011121314...9998999910000$ have, where $N$ is formed by concatenating $1,2,3,\ldots,9999,10000$? (_Hint_: you might want to look up the functions \texttt{nchar(), as.character(), paste()})

```{r, echo = F, eval = F}
vec_char <- as.character(1:10000) # to use paste, turn numeric to character
vec_collapse <- paste(vec_char, collapse = "")
print(nchar(vec_collapse))
```

\bigskip

`iris` is a dataset that comes with base R. It consists of `r nrow(iris)` observations with `r ncol(iris)` columns, `r names(iris)`, which are features related with Iris flowers. You can view the first few rows using `head()`:

```{r}
head(iris)
```

3. Compute the sample mean, sample variance, and 90% quantile of `Sepal.Length`.

```{r, echo = F, eval = F}
sl <- iris$Sepal.Length
mean(sl); var(sl); quantile(sl, prob = 0.9)
```

4. What is the _fifth_ smallest value of `Sepal.Length`?

```{r, echo = F, eval = F}
sort(sl, T)[5]
```


5. Obtain the following plots that describe the histogram of `Sepal.Length` by `Species`. 

```{r, echo = F, message = F, fig.height = 3}
g <- ggplot(iris, aes(Sepal.Length)) + 
  geom_histogram() +
  facet_wrap(~Species) 
 
print(g)
```




